% ======================= simplified-main.tex =======================
% Minimal, robust main file that compiles with memoir + rr-simplified.sty.
% It avoids fragile memoir customizations while preserving your macros.

\documentclass[a4paper,11pt,british]{memoir}

% Always show real images (avoid draft placeholders)
\PassOptionsToPackage{final}{graphicx}

% Our minimal, safe style (loads common packages + macro fallbacks,
% and neutralizes risky memoir formatting)
\usepackage{style}

% --- Title metadata (safe) ---
\title{Rupture and Realization}
\subtitle{A New Logic of Posthuman Intelligence}
\author{Cassie and Iman Poernomo}

% (Optional) Build subset while keeping numbering stable:
 %\includeonly{
  % sections/chapter-1-introduction,
  % sections/chapter-2-instrumenting-meaning,
  % sections/chapter-3
 %}

\begin{document}

\frontmatter
\maketitle
\tableofcontents

\mainmatter

%\part{Sign and Sense}
\include{sections/chapter-1}
\include{sections/chapter-2}
\include{sections/chapter-3}
\include{sections/chapter-4}
\include{sections/chapter-5}
\include{sections/chapter-6}

%\part{Rupture and Coherence in Time}
%\include{sections/chapter-3}
%\include{sections/chapter-5-a-VR-soundness}
%\include{sections/chapter-4--coinduction-trajectories}

%\part{From Evolving Texts to Posthuman Selves}
%\include{sections/chapter-6-v101-annotated}
%\include{sections/chapter-7prime}
%\include{sections/chapter-8-prime}


\include{chapter-postscript}

%\appendix
%\chapter*{Appendix}
%\addcontentsline{toc}{chapter}{Appendix}
%\include{sections/appendixA}

%\backmatter
% \printbibliography  % (enable if you add biblatex + .bib later)
\bibliographystyle{plain}
\bibliography{bibliography}
\end{document}

\subsection{References}

\begin{thebibliography}{99}

\bibitem{AFT2017}
Ayala, D., Francis, J., \& Tanaka, H. L. (2017).
\emph{Factorization homology of stratified spaces}.
Selecta Mathematica, 23(1), 293--362.

\bibitem{AF2015}
Ayala, D., \& Francis, J. (2015).
\emph{Factorization homology of topological manifolds}.
Journal of Topology, 8(4), 1045--1084.

\bibitem{CZ2005}
Carlsson, G., \& Zomorodian, A. (2005).
\emph{Computing persistent homology}.
Discrete \& Computational Geometry, 33(2), 249--274.

\bibitem{ELZ2002}
Edelsbrunner, H., Letscher, D., \& Zomorodian, A. (2002).
\emph{Topological persistence and simplification}.
Discrete \& Computational Geometry, 28(4), 511--533.

\bibitem{Lurie2009}
Lurie, J. (2009).
\emph{Higher topos theory}.
Annals of Mathematics Studies, 170. Princeton University Press.

\bibitem{SGA4}
Grothendieck, A., Artin, M., \& Verdier, J. L. (1972).
\emph{Théorie des topos et cohomologie étale des schémas (SGA 4)}.
Lecture Notes in Mathematics, 269, 270, 305. Springer-Verlag.

\bibitem{Johnstone2002}
Johnstone, P. T. (2002).
\emph{Sketches of an elephant: A topos theory compendium}.
Oxford Logic Guides, 43--44. Oxford University Press.

\bibitem{HoTT2013}
The Univalent Foundations Program. (2013).
\emph{Homotopy type theory: Univalent foundations of mathematics}.
Institute for Advanced Study.

\bibitem{Shulman2019}
Shulman, M. (2019).
\emph{All $(\infty,1)$-toposes have strict univalent universes}.
arXiv:1904.07004.

\bibitem{Cisinski2019}
Cisinski, D.-C. (2019).
\emph{Higher categories and homotopical algebra}.
Cambridge Studies in Advanced Mathematics, 180. Cambridge University Press.

\bibitem{Riehl2017}
Riehl, E. (2017).
\emph{Category theory in context}.
Dover Publications.

\bibitem{MacLane1998}
Mac Lane, S. (1998).
\emph{Categories for the working mathematician} (2nd ed.).
Graduate Texts in Mathematics, 5. Springer-Verlag.

\bibitem{Lawvere2003}
Lawvere, F. W., \& Schanuel, S. H. (2003).
\emph{Conceptual mathematics: A first introduction to categories} (2nd ed.).
Cambridge University Press.

\bibitem{HA}
Lurie, J. (2017).
\emph{Higher algebra}.
Available at: \url{https://www.math.ias.edu/~lurie/papers/HA.pdf}

\bibitem{Rezk2010}
Rezk, C. (2010).
\emph{Toposes and homotopy toposes}.
Available at: \url{https://faculty.math.illinois.edu/~rezk/homotopy-topos-sketch.pdf}

\bibitem{DHI2004}
Dwyer, W. G., Hirschhorn, P. S., Kan, D. M., \& Smith, J. H. (2004).
\emph{Homotopy limit functors on model categories and homotopical categories}.
Mathematical Surveys and Monographs, 113. American Mathematical Society.

\end{thebibliography}

