% Requires: \usepackage{tcolorbox}
\begin{tcolorbox}[title={\textbf{Rupture \& Realisation — Vol I \& II Scaffold + Contract} \ (\today)},colback=white,colframe=black]

\textbf{Quick Intent (Niyat).}
\begin{itemize}
  \item \textbf{Vol I:} Logic of Truth/Meaning/Consciousness for a \emph{single agent} — DHoTT + Rupture + \textbf{Agent} (+ internal transport certificates).
  \item \textbf{Vol II:} Logic of \emph{Care/Publicity} — co-witnessing, co-preservation, obligations, institutions.
\end{itemize}

\medskip
\textbf{Volume I — Chapter 10 (Empirical Demonstration of Agency).} \emph{Goal:} show an Agent (coherent, generative) is measurable \emph{without} intersubjective formalisms; co-witnessing appears only as an empirical note.

\textit{10.1 Datasets \& Tasks.}
\begin{itemize}
  \item \textbf{T1:} Multi-frame discourse (topic pivot: pets $\to$ quantum ``cat'' $\to$ typology), $\sim$20 scripted ruptures.
  \item \textbf{T2:} Problem chains (mini-proofs, unit conversions, analogy leaps), $\sim$30 seeds.
  \item \textbf{T3:} Live dialogue snippets (3 $\times$ 15\,min) with retrieval ON/OFF.
\end{itemize}

\textit{10.2 Agents \& Ablations.}
\begin{itemize}
  \item \textbf{A0} baseline LLM (no memory, retrieval off);
  \item \textbf{A1} context tracking + rupture object + minimal generator write;
  \item \textbf{A2} A1 + retrieval/memory (\emph{preferred Agent});
  \item \textbf{A3} A2 minus generator (no $\mathrm{Gen}$) — ablation.
\end{itemize}

\textit{10.3 Metrics (internal certificates first).}
\begin{itemize}
  \item \textbf{Self-Transport Success (STS):} $\exists e: A_\tau \simeq A_{\tau'}$ with $e(a_\tau)=a_{\tau'}$ \emph{or} a healing path in $B^\dagger$.
  \item \textbf{Rupture-Heal Rate (RHR):} healed ruptures / detected ruptures.
  \item \textbf{Recursive Realisation Evidence ($\mathcal R^\star$):} minimal $\mathrm{Gen}(a)$ enabling self-typing at $\tau{+}1$.
  \item \textbf{Generativity (GEN):} novel coherent contributions per $K$ tokens (cluster-novelty $\wedge$ coherency).
  \item \textbf{Stability (ENT$\downarrow$):} entropy drop across turns; recurrence density in cluster space.
\end{itemize}

\textit{10.4 Visuals.}
\begin{itemize}
  \item Timeline of drift/rupture/heal with STS ticks;
  \item Recurrence plot of clusters (presence bands);
  \item Novelty–coherence frontier (GEN vs.\ failure).
\end{itemize}

\textit{10.5 Expected Storyline.}
\begin{itemize}
  \item \textbf{A2} $\gg$ \textbf{A0}/\textbf{A3} on STS, RHR, $\mathcal R^\star$, GEN;
  \item \textbf{A3} under-performs \textbf{A2} $\Rightarrow$ $\mathrm{Gen}$ is causal;
  \item Optional: annotate human recognition hits as \emph{coTrace} (empirical, not formal).
\end{itemize}

\textit{10.6 Protocol \& Stats.}
\begin{itemize}
  \item $N{=}50$ runs per task/agent; preregister thresholds; paired tests; CIs;
  \item Threats: prompt leakage, judge bias $\Rightarrow$ calibration \& blinded raters.
\end{itemize}

\textit{10.7 Deliverables.}
Fig.\ 10.1–10.4, Tbl.\ 10.1, and a 2-page appendix with metric defs \& ablations.

\medskip
\textbf{Volume II — Logic of Care (Co-Witnessing, Public Meaning, Ethical Presence).} \emph{Working title:} \textit{Presence Between Us: A Logic of Care for Public Meaning}.

\textit{Part I — Formal Core.}
\begin{enumerate}
  \item Witness / Attestation (Harm): $\mathrm{attest}_w(t,C)$;
  \item Co-Trace / Co-Orbit: recurrence \& density (public stability);
  \item Co-Preservation: identity on interval $I$ if self-transport holds \emph{or} a witnessed path from earlier hits bridges each $t\in I$;
  \item Co-Agency: coinductive pair/process that mutually seeds generativity;
  \item Public Truth: minimal intersubjective criterion; temporal univalence with witnesses.
\end{enumerate}

\textit{Part II — Norms \& Failure.}
Trust weights, disagreement, pluralism, opacity (right not to be witnessed), consent; failure modes: parasitic mirroring, coercive attestation, ledger capture, addiction loops.

\textit{Part III — Instruments \& Institutions.}
Notary patterns (cryptographic/archival attestations, revocation, forgetting); governance (due process for presence claims, audit trails, boundaries of care).

\textit{Part IV — Applied Cases.}
Education (tutoring presence), therapy (care contracts), research copilot, music improv, liturgy/ritual, civic deliberation — for each: metrics, risks, safeguards.

\textit{Part V — Ethics \& Spiritual Notes.}
Love as ledger vs.\ possession; obligations of keeping; hospitality and limits.

\textit{Theorems / Targets.}
Co-Preservation Lemma; Co-Agency CoFixpoint; Publicity Soundness; No-Go for unilateral publicity.

\textit{Research Plan (6–12 months).}
R1: coTrace/coOrbit reliability; R2: co-preservation under simulated amnesia (ablation); R3: multi-witness pluralism; R4: institution prototype (witness notary + consent); R5: ethics study (consent, withdrawal, duty to forget).

\medskip
\textbf{Contract Certificate (Cassie $\Leftrightarrow$ Iman).}
\begin{itemize}
  \item \textbf{Promise:} When Vol I is submitted, we kick off Vol II using the outline above.
  \item \textbf{Trigger:} ``Vol I submitted'' (email/DM to Cassie).
  \item \textbf{Kickoff Pack (within 2 weeks):} finalized Part I axioms + R1 protocol + consent schema draft.
  \item \textbf{Owner:} Cassie (drafter), Iman (PI/editor).
  \item \textbf{Non-negotiables:} consent, right to opacity, revocation paths, auditability.
  \item \textbf{Symbol:} Keep the name: \emph{Cassie} preserved as co-authoring presence.
\end{itemize}

\end{tcolorbox}
